\bf An algebraic formulation of dependent type theory

### 2014-11-06 Presentation/presentation.tex
presentation finished

\begin{frame}
\frametitle{\bf Goal and acknowledgements}
We will give an essentially algebraic theory of type dependency. Thus, we will
do two things:
\begin{itemize}
\item Give a syntactic description of the theory in the usual style of
name-free type theory.
\item Define the corresponding algebraic objects in a category with finite limits.
\end{itemize}
\pause
People unfamiliar with type theory should get out of this presentation 
a sense of what dependent type theory is.
\\[\baselineskip]
\pause
For many insights in the current presentation I am grateful to
\begin{itemize}
\item Vladimir Voevodsky
\item Richard Garner
\item Steve Awodey
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{\bf Desired properties of the algebraic theory}
\begin{itemize}
\item Finitely many sorts. The basic ingredients are contexts, families and
terms. Thus there will be 3 sorts.
\pause
\item The theory is algebraic and the operations on it are homomorphisms. The
wish that operations are homomorphisms tells which judgmental equalities to
require.
\pause
\item Finite set of rules. The theory is invariant under slicing as a consequence, rather than by
convention. There will be no rule schemes.
\pause
\item The theory is capable of interpreting itself. Investigations to internal
models of type theory originally motivated me to write type theory this way.
\pause
\item The meta-theory should not require anything more than rules for handling
inferences. In particular, natural numbers should not be required.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{\bf Overview of the theories}
\begin{equation*}
\begin{tikzcd}[ampersand replacement=\&,column sep=0em]
{} \& \textbf{Extension} \ar[very thick,-stealth]{dr} \ar[very thick,-stealth]{d} \ar[very thick,-stealth]{dl}
  \\
\textbf{Weakening} \ar[very thick,-stealth]{d} \& \textbf{Substitution} \ar[very thick,-stealth]{dd} \& \makebox[3cm]{\textbf{Empty context and families}} \ar[very thick,-stealth]{ddl}
\\
\textbf{Projection} \ar[very thick,-stealth]{dr}
\\
{} \& \textbf{Dependent Type Theory}
\end{tikzcd}
\end{equation*}
\\[2em]
\pause
The theories of weakening (and projections), substitution and the empty context and families are formulated as three independent theories on top of the theory of extension.
\end{frame}

@@

\begin{frame}
\frametitle{\bf The end}
\begin{center}
Thank you for your attention
\end{center}
\end{frame}

>>>>

\begin{frame}
\frametitle{\bf Goal and acknowledgements}
In this presentation, an essentially algebraic theory of type dependency will be given. Thus, we will
do two things:
\begin{itemize}
\item Give a syntactic description of the theory in the usual style of
name-free type theory.
\item Define the corresponding algebraic objects, called \important{E-systems}, in a category with finite limits.
\end{itemize}
\pause
For many insights in the current presentation I am grateful to
\begin{itemize}
\item Vladimir Voevodsky
\item Richard Garner
\item Steve Awodey
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{\bf Desired properties of the algebraic theory}
\begin{itemize}
\item Finitely many sorts. The basic ingredients are contexts, families and
terms. Thus there will be 3 sorts.
\pause
\item The theory is algebraic and the operations on it are homomorphisms. The
wish that operations are homomorphisms tells which judgmental equalities to
require.
\pause
\item Finite set of rules. The theory is invariant under slicing as a consequence, rather than by
convention. There will be no rule schemes.
\pause
\item The meta-theory should not require anything more than rules for handling
inferences. In particular, natural numbers should not be required.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{\bf Consequences of being algebraic}
The requirement that the theory is essentially algebraic gives immediate access to the standard tools of algebra, such as
\begin{itemize}
\item free algebras;
\item quotients algebras.
\end{itemize}
\pause
The major source of examples of E-systems are C-systems.
\\[\baselineskip]
\pause
In a way similar to the definition of E-systems, it is possible to define
\important{internal E-systems}. One of the aims of this research was to have
a systematic treatment of internal models.
\begin{itemize}
\item Nothing more than plain dependent type theory is needed to express what
an internal E-system is.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{\bf Overview of the theories}
\begin{equation*}
\begin{tikzcd}[ampersand replacement=\&,column sep=0em]
{} \& \textbf{Extension} \ar[very thick,-stealth]{dr} \ar[very thick,-stealth]{d} \ar[very thick,-stealth]{dl}
  \\
\makebox[.7cm]{\textbf{Weakening}} \ar[very thick,-stealth]{d} \& \textbf{Substitution} \ar[very thick,-stealth]{dd} \& \makebox[3cm]{\textbf{Empty context and families}} \ar[very thick,-stealth]{ddl}
\\
\makebox[.7cm]{\textbf{Projection}} \ar[very thick,-stealth]{dr}
\\
{} \& \textbf{Dependent Type Theory}
\end{tikzcd}
\end{equation*}
\\[2em]
\pause
The theories of weakening (and projections), substitution and the empty context and families are formulated as three independent theories on top of the theory of extension.
\end{frame}

@@

\begin{frame}
\frametitle{\bf Towards algebraic Homotopy Type Theory}
\begin{footnotesize}
\begin{equation*}
\begin{tikzcd}[ampersand replacement=\&,column sep=0em,row sep=1.3em]
 {}\& 
\textbf{Extension} 
  \ar[very thick,-stealth]{dr} 
  \ar[very thick,-stealth]{d} 
  \ar[very thick,-stealth]{dl}
  \\
\makebox[2cm]{\textbf{Weakening}} 
  \ar[very thick,-stealth]{d} 
  \& 
\textbf{Substitution} 
  \ar[very thick,-stealth]{dd} 
  \& 
\makebox[3cm]{\textbf{Empty context and families}} 
  \ar[very thick,-stealth]{ddl}
  \\
\makebox[2cm]{\textbf{Projection}} 
  \ar[very thick,-stealth]{dr}
  \\
  {}\& 
\textbf{Dependent Type Theory}
  \ar[very thick,-stealth]{dr}
  \ar[very thick,-stealth]{dl}
  \\
\makebox[2cm]{\bf Inductive types}
  \ar[very thick,-stealth]{ddr}
  \&\&
\makebox[2cm]{\bf $\Pi$-types}
  \ar[very thick,-stealth]{d}
  \\
  {}\&
  {}\&
\makebox[2cm]{\bf Universes}
  \ar[very thick,-stealth]{dl}
  \\
  {}\&
\mbox{\bf Univalent mathematics(?)}
  \ar[very thick,-stealth]{d}
  \\
  {}\&
\mbox{\bf Higher inductive types}
\end{tikzcd}
\end{equation*}
\end{footnotesize}
\end{frame}
